---
title: "Summary of Assante Open Account"
author: "Arthur Ryman"
date: '2017-04-18'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

```{r date.of.death}
date.of.death <- as.Date("2016-07-25")
```

This document summarizes the transactions in Bella Ryman's Assante Open account `101947455NO`
for use in filing her 2016 personal income tax return.
Bella died on ``r date.of.death``.
I need to compute the unrealized capital gain (or loss) on the investments in this account because that amount is taxable.
I also need to compute the realized capital gain on all redemptions, the amount of dividends received, and the amount paid in management fees
as of the date of death.

## Data Preparation

```{r file.names}
txn.export.filename <- "2017-04-18-assante-transactions-export-bella-open.csv"
txn.table.file <- "2017-04-18-assante-transactions-table-bella-open"
txn.table.filename <- paste0(txn.table.file,".csv")
```
To prepare the data for this analysis, I did the following:

1. in Quicken
    1. enter transaction data from Bella's Assante statements
    2. view the transactions sorted by date
    3. export the transactions as CSV to ``r txn.export.filename``
2. in Excel
    1. edit the exported CSV file in Excel to turn it into a flat data table
    2. save the result as CSV to ``r txn.table.filename``

The following table lists the first few rows of the raw transaction data:
```{r read.csv}
txn <- read.csv(txn.table.filename, colClasses = "character")
knitr::kable(txn[1:6,], caption = "Sample of raw transaction data")
```

I read the CSV file as character data to avoid any automatic data conversions.
I explicitly convert the character data to its proper type in the following steps.

### Date

The `Date` column contains the transaction date in `YYYY-MM-DD` format.
It is convenient for income tax purposes to extract the year of the transaction
and to indicate if the transaction applies to the estate.
```{r date}
txn$Date <- as.Date(txn$Date)
txn$Year <- format(txn$Date, "%Y")
txn$Estate <- txn$Date > date.of.death
```

### Type

The `Type` column contains the transaction type.
```{r transaction.type}
txn$Type <- as.factor(txn$Type)
levels(txn$Type)
```

Transactions of type `Payment/Deposit` move cash into and out of the account.
They do not affect the the shares, so remove them.
```{r remove.payment.deposit}
txn <- txn[txn$Type != "Payment/Deposit",]
```

### Security

The `Security.Payee` column contains the name of the security or payee.
This column now only contains security names since I previously removed the `Payment/Deposit` transactions so it duplicates the `Security` column.
The `Security` column contains the security name.
```{r security}
txn$Security <- as.factor(txn$Security)
securities <- levels(txn$Security)
securities
```

### Price

The `Description` column describes the transaction, e.g. `5,747.296 shares @ 10.17`.
Quicken computes the price from the number of shares and the transaction amount, excluding other transaction costs such as commissions.
Assante reports prices to 4 significant digits but the last 2 appear to always be 0, so round off to 2 significant digits.
Extract the share price and store it in a new column named `Price`.
```{r price}
txn$Description <- gsub(",", "", txn$Description)
txn$Price <- round(as.numeric(sub("^.* shares @ ","",txn$Description)), digits = 2)
```

### Shares and Share.Balance

The `Shares` column contains the number of shares added to the holding for a security.
It is positive for `Buy` transactions and negative for `Sell` transactions.
The `Share.Balance` column contains the cummulative number of shares held for a security.
Quicken exports these values using a numeric format that uses a comma to separate thousands.
Remove the comma and convert these columns to numeric values.
`Shares` is exported with full accuracy but `Share.Balance` is sometimes rounded off to 2 significant digits.
For greater accuracy in reporting, use the value `Share.Balance.After` which is calculated below as the cummulative sum of `Shares`.
```{r shares}
txn$Shares <- as.numeric(gsub(",", "", txn$Shares))
txn$Share.Balance <- as.numeric(gsub(",", "", txn$Share.Balance))
```

### Amount

The `Amount` column contains the dollar amount by which the cash balance of the account changes in the transaction.
```{r amount}
txn$Amount <- as.numeric(gsub(",", "", txn$Amount))
```

Quicken reports `Amount` as 0 for dividend reinvestments because no cash enter or leaves the account.
The transaction costs for dividend reinvestments are 0.
Calculate the transaction amount for dividend reinvestments from the number of shares and the share price.
This quantity is reported in cents so round it off to 2 significant digits.
$$
Amount_{\mathrm{div}} = -Shares*Price
$$
```{r reinvest.dividend.amount}
div <- txn$Type == "Reinvest Dividend"
txn[div, "Amount"] <- -round(txn[div,"Shares"]*txn[div,"Price"], digits = 2)
```

### Txn.Costs

The transaction amount includes transaction costs such as commissions.
The transaction amount is positive for sales and negative for purchases.
Transaction costs are always non-negative and are reported to the nearest cent.
The transaction amount, number of shares added, share price, and transaction costs are related as follows:
$$
Amount + Shares * Price + Txn.Costs = 0
$$
Use this equation to calculate the transaction costs from the other quantities:
$$
Txn.Costs = -(Amount+Shares*Price)
$$
```{r txn.costs}

txn$Txn.Costs <- -round(txn$Amount + txn$Shares * txn$Price, digits = 2)

```
A transaction cost less than one dollar is rounding error.
Eliminate those. 

```{r}
txn$Txn.Costs <- ifelse(txn$Txn.Costs < 1, 0, txn$Txn.Costs)
```


The calculations of share balance, cost basis, etc. require that the transactions are sorted chronologically by date.
There may be many transactions on any given date.
It is therefore useful to assign a unique identifier to each transaction.
Sequence number is a suitable identifier for each transaction.
```{r seq}
txn$Seq <- 1:(dim(txn)[1])
```

Retain only the columns needed for the subsequent calculations.
```{r retain.needed.column}
txn <- txn[, c("Seq","Date", "Year", "Estate", "Type", "Security", "Price", "Shares", "Share.Balance", "Amount","Txn.Costs")]
knitr::kable(txn[1:6,], caption = "Sample of prepared transaction data", row.names = FALSE)
```

## Calculation of Cost Basis and Capital Gain

### Security Balances

At any given point in the transaction history, each security held in the account has a balance consisting of
the number of shares held and the cost basis in dollars of these shares.
The number of shares is always non-negative.
The cost basis is always non-negative.
Initially, both of these numbers are 0.
As transactions are made, the balance changes.
When the number of shares held is positive we can compute the average price paid for the shares.
For documentation purposes, we can also keep track of the date, sequence number, and type of the last transaction applied to the balance.

```{r balance.init}
balance <- data.frame(
  Security = securities, 
  Seq = 0, 
  Date = as.Date("1970-01-01"), 
  Type = "Buy", 
  Share.Balance = 0, 
  Cost.Basis = 0, 
  Avg.Price = 0, 
  row.names = "Security", 
  stringsAsFactors = FALSE)
knitr::kable(balance, caption = "Initial security balances")
```

### Processing the Transactions

Process each row of the transaction table in chronological order as follows.
Extend the transactions table with the following columns:

* security balances immediately before the transaction is made
    * Share.Balance.Before
    * Cost.Basis.Before
    * Avg.Price.Before = Cost.Basis.Before / Share.Balance.Before
* for purchases
    * Cost.of.Shares.Purchased = Shares \* Price = -(Amount + Txn.Costs)
* for sales
    * Cost.of.Shares.Sold = -Shares \* Avg.Price.Before
    * Proceeds.from.Shares.Sold = -Shares \* Price = Amount + Txn.Costs
    * Capital.Gain = Proceeds.from.Shares.Sold - Cost.of.Shares.Sold - Txn.Costs = Amount - Cost.of.Shares.Sold
* security balances immediately after the transaction is made
    * Share.Balance.After = Share.Balance.Before + Shares
    * Cost.Basis.After
        * for sales: Cost.Basis.Before - Cost.of.Shares.Sold
        * for purchases: Cost.Basis.Before + Cost.of.Shares.Purchased + Txn.Costs
    * Avg.Price.After = Cost.Basis.After / Share.Balance.After

#### Purchases

A purchase must increase the number of shares.
$$
Shares > 0
$$
A purchase must decrease the cash balance of the account.
$$
Amount < 0
$$

#### Sales

A sale must decrease the number of shares.
$$
Shares < 0
$$
The account must have enough shares to cover the sale.
$$
Share.Balance.After \geq 0
$$

##### Theorem

A sale that leaves a positive share balance does not change the average share price.

##### Proof

$$
\begin{aligned}
Avg.Price.After  &= \frac{Cost.Basis.After}{Share.Balance.After} \\
        &\\
        &= \frac{Cost.Basis.Before + Shares * Avg.Price.Before}{Share.Balance.Before + Shares} \\
        &\\
        &= \frac{Share.Balance.Before * Avg.Price.Before + Shares * Avg.Price.Before}{Share.Balance.Before + Share} \\
        &\\
        &= Avg.Price.Before
\end{aligned}
$$

###### QED

```{r process.txn}

txn$Share.Balance.Before <- txn$Cost.Basis.Before <- txn$Avg.Price.Before <- 0
txn$Cost.of.Shares.Purchased <- txn$Cost.of.Shares.Sold <- txn$Proceeds.from.Shares.Sold <- txn$Capital.Gain <- 0
txn$Share.Balance.After <- txn$Cost.Basis.After <- txn$Avg.Price.After <- 0

assertTrue <- function(condition, message, object) {
  if (!condition) {
    print(paste("Error:", message))
    print(object)
  }
}

for (i in 1:dim(txn)[1]) {

  s <- as.character(txn[i, "Security"])

  # copy the "before" security balances into the transaction
  
  txn[i, "Share.Balance.Before"] <- balance[s, "Share.Balance"]
  txn[i, "Cost.Basis.Before"] <- balance[s, "Cost.Basis"]
  txn[i, "Avg.Price.Before"] <- balance[s, "Avg.Price"]
  
  # apply the transaction
  txn[i, "Share.Balance.After"] <- txn[i, "Share.Balance.Before"] + txn[i, "Shares"]
  
  type <- as.character(txn[i, "Type"])
  if (type == "Sell") {
    
    assertTrue(txn[i, "Shares"] < 0, "a sell sale descrease the number of shares", txn[i,])
    assertTrue(txn[i, "Share.Balance.After"] >= 0, "the account must have enough shares to cover the sale", txn[i,])

    txn[i, "Proceeds.from.Shares.Sold"] <- round(txn[i, "Amount"] + txn[i, "Txn.Costs"], 2)
    txn[i, "Cost.of.Shares.Sold"] <- round(-txn[i, "Shares"] * txn[i, "Avg.Price.Before"], 2)
    txn[i, "Capital.Gain"] <- round(txn[i, "Proceeds.from.Shares.Sold"] - txn[i, "Cost.of.Shares.Sold"] - txn[i, "Txn.Costs"], 2)
    txn[i, "Cost.Basis.After"] <- round(txn[i, "Cost.Basis.Before"] - txn[i, "Cost.of.Shares.Sold"], 2)

  } else {

    # Purchase
    
    assertTrue(txn[i, "Shares"] > 0, "a purchase must increase the number of shares", txn[i,])
    assertTrue(txn[i, "Amount"] < 0, "a purchase must descrease the cash balance of the account", txn[i,])

    txn[i, "Cost.of.Shares.Purchased"] <- -round(txn[i, "Amount"] + txn[i, "Txn.Costs"], 2)
    txn[i, "Cost.Basis.After"] <- round(txn[i, "Cost.Basis.Before"] + txn[i, "Cost.of.Shares.Purchased"] + txn[i, "Txn.Costs"], 2)
  }
  
  txn[i, "Avg.Price.After"] <- ifelse(txn[i, "Share.Balance.After"] > 0, round(txn[i, "Cost.Basis.After"] / txn[i, "Share.Balance.After"], digits = 4), 0)
  
  # update the security balance
  balance[s, "Seq"] <- txn[i, "Seq"]
  balance[s, "Date"] <- txn[i, "Date"]
  balance[s, "Type"] <- type
  balance[s, "Share.Balance"] <- txn[i, "Share.Balance.After"]
  balance[s, "Cost.Basis"] <- txn[i, "Cost.Basis.After"]
  balance[s, "Avg.Price"] <- txn[i, "Avg.Price.After"]
}

knitr::kable(balance, caption = "Updated account balances")
```



The market value of a security resulting from the transaction is the share balance times the share price.
Store the market value in a new column named `Market.Value`.
```{r market.value}
txn$Market.Value <- round(txn$Share.Balance.After * txn$Price, 2)
```

```{r calculated.cost.basis}
knitr::kable(txn[1:6,], caption = "Sample of calculated cost basis, etc.", row.names = FALSE)
```

## Calculate the Unrealized Capital Gain at the Date of Death

Find the most recent transaction for each security up to the date of death since that will contain the final cost basis for the security.
```{r final.cost.basis}
max.Seq <- rep_len(0, length(securities))
for (i in 1:length(securities)) {
  max.Seq[i] <- max(txn[txn$Security == securities[i] & txn$Date <= date.of.death, "Seq"])
}
final.txn <- txn[max.Seq, c("Security", "Share.Balance", "Cost.Basis.After")]
final.txn <- final.txn[final.txn$Share.Balance > 0,]
knitr::kable(final.txn, caption = "Final Transactions", row.names = FALSE)
```

The share prices on the date of death are as follows:
```{r final.share.price}
final.share.price <- read.csv("2016-07-25-share-price.csv", colClasses = "character")
final.share.price$Price <- as.numeric(final.share.price$Price)
knitr::kable(final.share.price, row.names = FALSE, caption = "Share price on 2016-07-25")
```

Merge the final price into the final transaction table and compute the market value and capital gain.
```{r merge.price}
final.txn$Price <- 0
for (s in final.share.price$Security) {
  final.txn[final.txn$Security == s, "Price"] <- final.share.price[final.share.price$Security == s, "Price"]
}
final.txn$Market.Value <- round(final.txn$Share.Balance * final.txn$Price, 2)
final.txn$Capital.Gain <- round(final.txn$Market.Value - final.txn$Cost.Basis.After, 2)
knitr::kable(final.txn, caption = "Final Transactions", row.names = FALSE)
```

Summarize the capital gain.
```{r summarize.captial.gain}
summary.final.txn <- data.frame(
  Cost.Basis.After = sum(final.txn$Cost.Basis.After), 
  Market.Value = sum(final.txn$Market.Value),
  Capital.Gain = sum(final.txn$Capital.Gain))
knitr::kable(summary.final.txn, caption = "Summary of capital gain")
```


```{r write.csv}
final.file <- paste0(txn.table.file,"-final.csv")
write.csv(final.txn, file = final.file, row.names = FALSE)

output.file <- paste0(txn.table.file,"-out.csv")
write.csv(txn, file = output.file, row.names = FALSE)
```

## Appendix

This appendix lists the raw data.

### Share Prices

### Transactions

```{r txn}
knitr::kable(txn[1:10,])
```